Management induced changes of soil organic carbon on global croplands

### 2020-12-11 SOCbudgetPaper/methods.Rmd
some smaller updates in method

In a simple first order kinetic approach the steady-state soil organic carbon stocks $SOC^{\mathrm{eq}}$ are given by 
\begin{equation}
SOC^{\mathrm{eq}}_{i,t,sub,lu} =\frac{C^{\mathrm{in}}_{i,t,sub,lu}}{k_{i,t,sub,lu}}
(\#eq:inoutflow)
\end{equation}
with $C^{\textrm{in}}$ being the carbon inputs to the soil, $k$ denotes the soil organic carbon decay rate. This equation is valid for all grid cells $i$ and all years $t$. We use the steady-state method for our calculations, which assumes three soil carbon sub-pools $sub$ (active, slow and passive) and interactions between them, following the approach in the Century model (@parton_analysis_1987). Annual carbon inflow to each sub-pool and annual decay rates of each sub-pool are land-use type $lu$ specific. 
We distinguish two land-use types: cropland and uncropped land under potential natural vegetation as representative for all other land-use types including forestry and pastures (referred to as natural vegetation in the following).

Carbon inputs for cropland are below and above ground residues left or returned to the field (see Sect. \@ref(sec:residues)) and manure inputs (see Sect. \@ref(sec:livstmanure)); for natural vegetation litterfall (including fine root turnover) ([@schaphoff_lpjml4_2018]) is the only source of carbon inflow to the soil. Following the IPCC guidelines (@calvo_buendia_ipcc_2019), carbon inputs are disaggregated into metabolic and structural components depending on their lignin and nitrogen content. For each component the sum of all carbon input sources is allocated to the respective $SOC$ sub-pools via transfer coefficients. This implies that both the amount of carbon and its structural composition determine the effective inflow into the different pools. Data sources for all considered carbon inputs as well as for lignin and nitrogen content are listed in Table \@ref(tab:datasourceinputs).

@@

According to the IPCC guidelines $SOC$ changes can be expressed as the difference of two consecutive years (see Eq. 5.0A in [@calvo_buendia_ipcc_2019]). This, however, will also include naturally occurring changes due to climatic varition over time. For our study, we defined the absolute and relative SOC changes in relation to a potential natural state $SOC^{\mathrm{pnv}}$ under the same climatic conditions in grid cell $i$ at time $t$, that is based on the natural vegetation $SOC$ calculations as defined above without accounting for land conversion from cropland at any time. The absolute changes $\Delta SOC$ and relative changes $F^{\mathrm{SCF}}$ are thus given by

@@

Land-use patterns are based on the Land-Use Harmonization 2 [@hurtt_harmonization_2020] data set, which we sum up from quarter-degree to half-degree resolution. We disaggregate the physical area of the five different cropland subcategories (c3ann, c3per, c4ann, c4per, c3nfx) of LUH2 into our 17 crop groups, applying the relative shares for each grid cell based on the country- and year-specific area harvested shares of FAOSTAT data [@faostat_faostat_2016] (see Table \@ref(append:Tableluh2fao2mag) for more details on the crop group mapping). By calculating country-specific multicropping factors using FAOSTAT data, we are able to compute crop-group specific area harvested on grid cell level. 
Land-use transitions are calculated as net area differences of the land-use data at half-degree resolution, considering no split up into crop-group specific areas but only total cropland and natural vegetation areas.

@@

From our 17 crop groups, we compile four residue groups (straw, high- and low-lignin residues, residues without dual use), of which the first three are taken away from the field for other purposes (see Table \@ref(append:Tablekcr2kres)). Residue feed demand for five different livestock groups is based on country-specific feed baskets (see [@weindl_livestock_2017]), that differentiate between the residue groups and take available $AGR$ biomass as well as livestock productivity into account. We estimate a material-use share for the straw-residue group of 5% and a fuel-share of 10% for all used residue groups in low-income countries. For high-income countries, no withdrawal for material or fuel use is assumed, and use shares of middle-income countries are linearly interpolated based on per-capita income, following the same rationale as for the share of burnt residues described above. The remaining $AGR$ as well as all $BGR$ are expected to be left on the field. We limit high returning rates at $10\tfrac{\unit{tC}}{\unit{ha}}$ to correct for outliers.

@@

Manure production and management is based on a revised methodology of [@bodirsky_n2o_2012] and is presented here due to its central role in soil carbon modeling. Based on the gridded livestock distribution we calculate spatially explicit excretion by estimating the nitrogen balance of the livestock systems on the basis of comprehensive livestock feed baskets [@weindl_livestock_2017], assuming that all nitrogen in protein feed intake, minus the nitrogen in the slaughter mass, is excreted. Carbon in excreted manure is estimated by applying fixed c/n ratios (given by [@calvo_buendia_ipcc_2019]). 
Depending on the feed system we assume manure to be handled in four different ways:
All manure originated from pasture feed intake is excreted directly on pastures and rangelands (pasture grazing), deducting manure collected as fuel.
Whereas for low-income countries, we adopt a share of 25% of crop residues in feed intake directly consumed and excreted on crop fields (stubble grazing), we do not consider any stubble grazing in high-income countries; middle-income countries see linearly interpolated shares depending on their per-capita income.
For all other feed items, we assume the manure to be stored in animal waste management systems associated with livestock housing. 
To estimate the carbon actually returned to the soil, we account for carbon losses during storage, where return shares depend on different animal waste management and grazing systems. Whereas we assume no losses for pasture and stubble grazing, we consider that the manure collected as fuel is not returned to the fields. For manure stored in different animal waste management systems we compiled carbon loss rates depending on the different systems and the associated nitrogen loss rates as specified in [@bodirsky_n2o_2012] (see Sect. \@ref(append:TableclossAWMS) for more details). We limit high application shares at $10\tfrac{\unit{tC}}{\unit{ha}}$ to correct for outliers.

@@

In order to derive a spatial distribution of the three different tillage types specified by the IPCC -- full tillage, reduced tillage and no tillage --, we assume that all natural land and pastures are not tilled, whereas annual crops are under full and perennials under reduced tillage per default. Furthermore, we assume no tillage in cropland cells specified as no tillage cell based on the historic global gridded tillage data set from  [@porwollik_generating_2018]. This data set is extended to the period of 1975-2010 by combining country-level data on areas under conservation agriculture from [@fao_aquastat_2016] available from 1975 on with LUH2 physical crop areas, applying the methodology of [@porwollik_generating_2018] to identify potential no-tillage grid-cells.

@@

Due to missing information about the litterfall composition, we assume woody potential natural vegetation for the lignin-to-carbon (short lg/c) and nitrogen-to-carbon (short n/c) ratios for all land under natural vegetation. This is however a strong simplification. Therefore, we test the sensitivity of the model for other natural vegetation parameterizations. Perennial grasses (scenario name \textit{LitterPNV-PerennialGrasses}) are given by the IPCC guidelines (\cite{calvo_buendia_ipcc_2019}), but for woody biomass we used the CENTURY configuration file ([@century_model_2000]) for trees as parameterization were not given. Whereas lg/c ratios ranges between $0.09-0.35$ with an average of $0.24$ (default value) for different tree compartments and different tree types, n/c ratios range between $0.001-0.05$ with an average of $0.015$ (default value). We decided to average over all tree compartments and tree types equally for the default values, since we have no data available on tree type distribution and tree compartment composition. We test however a range of different plant types as specified in Table @ref(tab:scen_litterpnv).

@@

 \begin{table*}[h]
 \caption{Scenarios for sensitivity analysis on litter parameterization}
 \begin{tabular}{l l l}
 \tophline
  Scenario name & plant type & n/c & lg/c \\
  \middlehline
  \textit{LitterPNV-CenturyAverage} & mix over all tree types & 0.015  & 0.25 \\
  \textit{LitterPNV-PerennialGrasses} & perennial grasses     & 0.029  & 0.11 \\
  \textit{LitterPNV-AASS}  & Arctic alpine subshrub           & 0.022  & 0.22 \\
  \textit{LitterPNV-BLEAF} & Temperate broadleaf evergreen    & 0.013  & 0.27 \\
  \textit{LitterPNV-BRCF}  & Boreal coniferous forest         & 0.013  & 0.17 \\
  \textit{LitterPNV-MTCF}  & Maritime coniferous forest       & 0.0094 & 0.22 \\
  \textit{LitterPNV-TREF}  & Tropical evergreen forest        & 0.016  & 0.30 \\
 \bottomhline
 \end{tabular}
 \belowtable{}
 (\#tab:scen_litterpnv)
 \end{table*}
```

>>>>

In a simple first order kinetic approach the steady-state soil organic carbon stocks $SOC^{\mathrm{eq}}$ are given by 
\begin{equation}
SOC^{\mathrm{eq}}_{i,t,sub,lu} =\frac{C^{\mathrm{in}}_{i,t,sub,lu}}{k_{i,t,sub,lu}}
(\#eq:inoutflow)
\end{equation}
with $C^{\textrm{in}}$ being the carbon inputs to the soil, $k$ denotes the soil organic carbon decay rate. This equation is valid for all grid cells $i$ and all years $t$. We use the steady-state method for our calculations, which assumes three soil carbon sub-pools $sub$ (active, slow and passive) and interactions between them, following the approach in the Century model [(@parton_analysis_1987]). Annual carbon inflow to each sub-pool and annual decay rates of each sub-pool are land-use type $lu$ specific. 
We distinguish two land-use types: cropland and uncropped land under potential natural vegetation as representative for all other land-use types including forestry and pastures (referred to as natural vegetation in the following).

Carbon inputs for cropland are below and above ground residues left or returned to the field (see Sect. \@ref(sec:residues)) and manure inputs (see Sect. \@ref(sec:livstmanure)); for natural vegetation litterfall (including fine root turnover) ([@schaphoff_lpjml4_2018]) is the only source of carbon inflow to the soil. Following the IPCC guidelines [@calvo_buendia_ipcc_2019], carbon inputs are disaggregated into metabolic and structural components depending on their lignin and nitrogen content. For each component the sum of all carbon input sources is allocated to the respective $SOC$ sub-pools via transfer coefficients. This implies that both the amount of carbon and its structural composition determine the effective inflow into the different pools. Data sources for all considered carbon inputs as well as for lignin and nitrogen content are listed in Table \@ref(tab:datasourceinputs).

@@

According to the IPCC guidelines $SOC$ changes can be expressed as the difference of two consecutive years (see Eq. 5.0A in [@calvo_buendia_ipcc_2019]). This, however, will also include naturally occurring changes due to climatic variation over time. For our study, we defined the absolute and relative SOC changes in relation to a potential natural state $SOC^{\mathrm{pnv}}$ under the same climatic conditions in grid cell $i$ at time $t$, that is based on the natural vegetation $SOC$ calculations as defined above without accounting for land conversion from cropland at any time. The absolute changes $\Delta SOC$ and relative changes $F^{\mathrm{SCF}}$ are thus given by

@@

Land-use patterns are based on the Land-Use Harmonization 2 [@hurtt_harmonization_2020] data set, which we sum up from quarter-degree to half-degree resolution. We disaggregate the physical area of the five different cropland subcategories (c3ann, c3per, c4ann, c4per, c3nfx) of LUH2 into our 17 crop groups, applying the relative shares for each grid cell based on the country- and year-specific area harvested shares of FAOSTAT data [@faostat_faostat_2016] (see FAO2LUH2MAG_croptypes.csv in [@mrcommons]). By calculating country-specific multicropping factors using FAOSTAT data, we are able to compute crop-group specific area harvested on grid cell level. 
Land-use transitions are calculated as net area differences of the land-use data at half-degree resolution, considering no split up into crop-group specific areas but only total cropland and natural vegetation areas.

@@

From our 17 crop groups, we compile four residue groups (straw, high- and low-lignin residues, residues without dual use), of which the first three are taken away from the field for other purposes (see mappingCrop2Residue.csv in [@mrcommons]). Residue feed demand for five different livestock groups is based on country-specific feed baskets (see [@weindl_livestock_2017]), that differentiate between the residue groups and take available $AGR$ biomass as well as livestock productivity into account. We estimate a material-use share for the straw-residue group of 5% and a fuel-share of 10% for all used residue groups in low-income countries. For high-income countries, no withdrawal for material or fuel use is assumed, and use shares of middle-income countries are linearly interpolated based on per-capita income, following the same rationale as for the share of burnt residues described above. The remaining $AGR$ as well as all $BGR$ are expected to be left on the field. We limit high returning rates at $10\unit{tC}\unit{ha}^{-1}}$ to correct for outliers.

@@

Manure production and management is based on a revised methodology of [@bodirsky_n2o_2012] and is presented here due to its central role in soil carbon modeling. Based on the gridded livestock distribution we calculate spatially explicit excretion by estimating the nitrogen balance of the livestock systems on the basis of comprehensive livestock feed baskets [@weindl_livestock_2017], assuming that all nitrogen in protein feed intake, minus the nitrogen in the slaughter mass, is excreted. Carbon in excreted manure is estimated by applying fixed c/n ratios (given by [@calvo_buendia_ipcc_2019]). 
Depending on the feed system we assume manure to be handled in four different ways:
All manure originated from pasture feed intake is excreted directly on pastures and rangelands (pasture grazing), deducting manure collected as fuel.
Whereas for low-income countries, we adopt a share of 25% of crop residues in feed intake directly consumed and excreted on crop fields (stubble grazing), we do not consider any stubble grazing in high-income countries; middle-income countries see linearly interpolated shares depending on their per-capita income.
For all other feed items, we assume the manure to be stored in animal waste management systems associated with livestock housing. 
To estimate the carbon actually returned to the soil, we account for carbon losses during storage, where return shares depend on different animal waste management and grazing systems. Whereas we assume no losses for pasture and stubble grazing, we consider that the manure collected as fuel is not returned to the fields. For manure stored in different animal waste management systems we compiled carbon loss rates depending on the different systems and the associated nitrogen loss rates as specified in [@bodirsky_n2o_2012] (see Sect. \@ref(append:TableclossAWMS) for more details). We limit high application shares at $10\unit{tC}\unit{ha}^{-1}}$ to correct for outliers.

@@

In order to derive a spatial distribution of the three different tillage types specified by the IPCC -- full tillage, reduced tillage and no tillage --, we assume that all natural land and pastures are not tilled, whereas annual crops are under full and perennials under reduced tillage per default. Furthermore, we assume no tillage in cropland cells specified as no tillage cell based on the historic global gridded tillage data set from  [@porwollik_generating_2018]. This data set is extended to the period of 1975--2010 by combining country-level data on areas under conservation agriculture from [@fao_aquastat_2016] and half-degree resolution physical crop areas from [@hurtt_harmonization_2020], applying the methodology of [@porwollik_generating_2018] to identify potential no-tillage grid cells.

@@

Due to missing information about the litterfall composition, we assume woody potential natural vegetation for the lignin-to-carbon (short lg/c) and nitrogen-to-carbon (short n/c) ratios for all land under natural vegetation. This is however a strong simplification. Therefore, we test the sensitivity of the model for other natural vegetation parameterizations. Perennial grasses (scenario name \textit{LitterPNV-PerennialGrasses}) are given by the IPCC guidelines ([@calvo_buendia_ipcc_2019]), but for woody biomass we used the CENTURY configuration file ([@century_model_2000]) for trees as parameterization were not given. Whereas lg/c ratios ranges between $0.09-0.35$ with an average of $0.24$ (default value) for different tree compartments and different tree types, n/c ratios range between $0.001-0.05$ with an average of $0.015$ (default value). We decided to average over all tree compartments and tree types equally for the default values, since we have no data available on tree type distribution and tree compartment composition. We test however a range of different plant types as specified in Table @ref(tab:scen_litterpnv).

@@

 \begin{table*}[h]
 \caption{Scenarios for sensitivity analysis on litter parameterization}
 \begin{tabular}{l l l}
 \tophline
  Scenario name & plant type & n/c & lg/c \\
  \middlehline
  \textit{LitterPNV-Mean} & mix over all tree types & 0.015  & 0.25 \\
  \textit{LitterPNV-Grass} & perennial grasses     & 0.029  & 0.11 \\
  \textit{LitterPNV-AASS}  & Arctic alpine subshrub           & 0.022  & 0.22 \\
  \textit{LitterPNV-BLEAF} & Temperate broadleaf evergreen    & 0.013  & 0.27 \\
  \textit{LitterPNV-BRCF}  & Boreal coniferous forest         & 0.013  & 0.17 \\
  \textit{LitterPNV-MTCF}  & Maritime coniferous forest       & 0.0094 & 0.22 \\
  \textit{LitterPNV-TREF}  & Tropical evergreen forest        & 0.016  & 0.30 \\
 \bottomhline
 \end{tabular}
 \belowtable{}
 (\#tab:scen_litterpnv)
 \end{table*}
```
